---
title: "HW2 P8105"
author: "Lauren Littig"
date: "2026-10-02"
output: github_document
---

# Problem 1

Load packages, import/clean data, select needed variables, convert entry variable from character to logical:

```{r}
library(tidyverse)
library(readxl)
subway_df =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names()

names(subway_df)

subway_df =
  select(subway_df, line:entry, vending, ada) |>
  mutate(
    entry =
      case_match(
        entry,
        "YES" ~ TRUE,
        "NO" ~ FALSE
      )
  )
```

`subway_df` contains the variables `r paste(names(subway_df))`. To get my dataframe into the current format, I first cleaned the data, selected needed variables, and converted the entry variable from character to logical. The current dataframe contains `r nrow(subway_df)` observations and `r ncol(subway_df)` variables. These data are NOT tidy because route is spread across many variables.

How many distinct stations are there?

```{r}
subway_df |>
  distinct(station_name, line) |>
  nrow()
```

There are 465 distinct stations.

How many stations are ADA compliant?

```{r}
subway_df |> 
  filter(ada == TRUE) |>
  distinct(station_name, line) |>
  nrow()
```

84 stations are ADA compliant.

### What proportion of station entrances / exits without vending allow entrance?

```{r}
subway_df |>
  filter(vending == "NO") |> 
  summarize(
    proportion_entry = mean(entry, na.rm = TRUE)
  )
```

37.7% of station entrances / exits without vending allow entrance.

### Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?

```{r}
subway_df =
  subway_df |> 
  mutate(
    across(route1:route11, as.character)
  ) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name")

subway_df |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |> 
  nrow()

# 60 distinct stations serve the A train. 

subway_df |> 
  filter(route_name == "A", ada == TRUE) |> 
  distinct(line, station_name) |> 
  nrow()

# 17 distinct stations serve the A train and are ADA compliant.
```

# Problem 2

```{r}
trash_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx", 
    sheet = 1,
    range = "A2:N742"
    ) |> 
  janitor::clean_names() |> 
  mutate(wheel = "Mr",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles),
         sports_balls = as.integer(round(sports_balls))
  )
```

Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda, and combine this with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining.

```{r}
trash_df_2 =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx", 
    sheet = 2,
    range = "A2:M139"
    ) |> 
  janitor::clean_names() |> 
  mutate(
    wheel = "Prof",
    plastic_bottles = as.numeric(plastic_bottles),
    year = as.numeric(year)
  )

trash_df_3 =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx", 
    sheet = 4,
    range = "A2:L394"
    ) |> 
  janitor::clean_names() |> 
  mutate(
    wheel = "Gwynnda",
    plastic_bottles = as.numeric(plastic_bottles),
    year = as.numeric(year)
  )
```

# Combine Trash Wheel Data Sets

```{r}
trash_df_full =
  bind_rows(trash_df, trash_df_2, trash_df_3) |> 
  relocate(wheel)
```

The combined `trash_df_full` dataset contains `r nrow(trash_df_full)` observations and `r ncol(trash_df_full)` variables. Each observation represents a dumpster collected by one of the three Trash Wheels. Key variables include `r paste(names(trash_df_full))`.

The total weight of trash collected by Professor Trash Wheel is `r trash_df_full |> 
filter(wheel == "Prof") |> 
pull(weight_tons) |> 
sum(na.rm = TRUE)` tons.

The total number of cigarette butts collected by Gwynnda in June of 2022 was `r trash_df_full |> 
filter(wheel == "Gwynnda", month == "June", year == 2022) |> 
pull(cigarette_butts) |> 
sum(na.rm = TRUE)`.

# Problem 3

Import, clean, and tidy the dataset of baseline demographics. Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric), and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline). Discuss important steps in the import process and relevant features of the dataset. How many participants were recruited, and of these how many develop MCI? What is the average baseline age? What proportion of women in the study are APOE4 carriers?

```{r}
mci_baseline_df =
  read_csv("data/MCI_baseline.csv", skip = 1, na = c("", ".", "NA")) |> 
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex, 
      1 ~ "Male",
      0 ~ "Female"
      ),
    apoe4 = case_match(
      apoe4,
      1 ~ "Carrier",
      0 ~ "Non-carrier"
    )
  ) |> 
  filter(is.na(age_at_onset) | age_at_onset > current_age)

```

To get my dataframe to its current form, important steps included importing the dataset and skipping line 1 which contains information about the dataset, as well as telling R to treat "." as NA. I also recoded my numeric `sex` and `APOE4` variables into descriptive categorical variables so they could be interpreted more easily.

`r nrow(mci_baseline_df)` eligible participants were recruited and `r mci_baseline_df |> filter(!is.na(age_at_onset)) |> nrow()` developed MCI during the follow-up period.

The average baseline age is `r mean(pull(mci_baseline_df, current_age))` and the proportion of women in the study who are APOE4 carriers is `r mci_baseline_df |> filter(sex == "Female") |> summarize(prop_APOE4 = mean(apoe4 == "Carrier")) |> pull(prop_APOE4)`.

# Import the Biomarkers dataset

```{r}
biomarkers_df = 
  read_csv("data/mci_amyloid.csv", skip = 1, na = c("", ".", "NA", "Na")) |> 
  janitor::clean_names() |> 
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    names_prefix = "time_",
    values_to = "amyloid_ratio"
  ) |> 
  mutate(
    time = replace(time, time == "baseline", 0),
    time = as.numeric(time),
    amyloid_ratio = as.numeric(amyloid_ratio)
  ) |> 
  rename(id = study_id)
```

I imported the dataset, defined NAs (including "Na" as this was used in the file), cleaned the names, pivoted the time columns to tidy my dataset, and made my `time` and `amyloid_ratio` variables numeric. This dataset contains `r nrow(biomarkers_df)` participant-time observations from `r n_distinct(pull(biomarkers_df, id))` participants and `r ncol(biomarkers_df)` variables: participant ID, time, and amyloid ratio.

### Check whether some participants appear only in the baseline or amyloid datasets:

```{r}
mci_baseline_df |> 
  anti_join(biomarkers_df, by = "id")
```

There are 8 participants in the baseline dataset that are not in the amyloid dataset.

```{r}
biomarkers_df |> 
  anti_join(mci_baseline_df, by = "id") |> 
  distinct(id) |> 
  nrow()
```

There are 16 unqiue participants in the amyloid dataset that are not in the baseline dataset.

### Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, and briefly describe the resulting dataset; export the result as a CSV to your data directory.

```{r}
combined_df =
  inner_join(mci_baseline_df, biomarkers_df, by = "id")
```

The combined dataset contains `r n_distinct(pull(combined_df, id))` participants, 
`r nrow(combined_df)` participant-time observations, and`r ncol(combined_df)` variables. Only participants with both baseline demographic information and amyloid biomarker data are included.

```{r}
write_csv(combined_df, "data/combined_df.csv")
```

