---
title: "p8105_hw2_LL3969"
author: "Lauren Littig"
date: "2026-10-02"
output: github_document
---

# Problem 1

Load packages, import/clean data, select needed variables, convert entry variable from character to logical:

```{r}
library(tidyverse)
subway_df =
  read_csv("NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names()

names(subway_df)

subway_df =
  select(subway_df, line:entry, vending, ada) |>
  mutate(
    entry =
      case_match(
        entry,
        "YES" ~ TRUE,
        "NO" ~ FALSE
      )
  )
```

`subway_df` contains the variables `r paste(names(subway_df))`. To get my dataframe into the current format, I first cleaned the data, selected needed variables, and converted the entry variable from character to logical. The current dataframe contains `r nrow(subway_df)` observations and `r ncol(subway_df)` variables. These data are NOT tidy because (something about route)....

How many distinct stations are there?

```{r}
subway_df |>
  distinct(station_name, line) |>
  nrow()
```

How many stations are ADA compliant?

```{r}
subway_df |> 
  filter(ada == TRUE) |>
  distinct(station_name, line) |>
  nrow()
```

84 stations are ADA compliant.

What proportion of station entrances / exits without vending allow entrance?

```{r}
subway_df |>
  filter(vending == "NO") |> 
  summarize(
    proportion_entry = mean(entry, na.rm = TRUE)
  )
```

37.7% of station entrances / exits without vending allow entrance.
